---
title: "Case Study - Real Time Location Systems"
date: "`r Sys.Date()`"
author: "Allison Roderick, Jenna Ford, William Arnost"
output:
  rmdformats::readthedown:
    code_folding: hide
    highlight: kate
    toc_depth: 3
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Data Ingest

```{r}
processLine = function(x){
  tokens = strsplit(x, "[;=,]")[[1]]
  
  if (length(tokens) == 10) 
    return(NULL)
 
  tmp = matrix(tokens[ - (1:10) ], ncol= 4, byrow = TRUE)
  cbind(matrix(tokens[c(2, 4, 6:8, 10)], nrow(tmp), 6, 
               byrow = TRUE), tmp)
}


roundOrientation = function(angles) {
  refs = seq(0, by = 45, length  = 9)
  q = sapply(angles, function(o) which.min(abs(o - refs)))
  c(refs[1:8], 0)[q]
}

readData = function(filename = 'offline.final.trace.txt', 
           subMacs = c("00:0f:a3:39:e1:c0", "00:0f:a3:39:dd:cd", "00:14:bf:b1:97:8a",
                       "00:14:bf:3b:c7:c6", "00:14:bf:b1:97:90", "00:14:bf:b1:97:8d",
                       "00:14:bf:b1:97:81"))  {
    txt = readLines(filename)
    lines = txt[ substr(txt, 1, 1) != "#" ]
    tmp = lapply(lines, processLine)
    offline = as.data.frame(do.call("rbind", tmp), 
                            stringsAsFactors= FALSE) 
    
    names(offline) = c("time", "scanMac", 
                       "posX", "posY", "posZ", "orientation", 
                       "mac", "signal", "channel", "type")
    
     # keep only signals from access points
    offline = offline[ offline$type == "3", ]
    
    # drop scanMac, posZ, channel, and type - no info in them
    dropVars = c("scanMac", "posZ", "channel", "type")
    offline = offline[ , !( names(offline) %in% dropVars ) ]
    
    # drop more unwanted access points
    offline = offline[ offline$mac %in% subMacs, ]
    
    # convert numeric values
    numVars = c("time", "posX", "posY", "orientation", "signal")
    offline[ numVars ] = lapply(offline[ numVars ], as.numeric)

    # convert time to POSIX
    offline$rawTime = offline$time
    offline$time = offline$time/1000
    class(offline$time) = c("POSIXt", "POSIXct")
    
    # round orientations to nearest 45
    offline$angle = roundOrientation(offline$orientation)
      
    return(offline)
  }
```
```{r}
offline = readData(filename = 'offline.final.trace.txt')
subMacs = names(sort(table(offline$mac), decreasing = TRUE))
str(offline)
```
```{r}
online = readData(filename = 'online.final.trace.txt')
str(online)
```

# Signal Stength Analysis

## Figure 1.6: Signal Stength by Angle for Each Access Point  
*The boxplots in this figure represent signals for one location which is in the upper left corner of the floor plan (x=2,y=12). These boxes are organized by access point and angle of the hand-held device. The dependence of signal strength on angle is apparent at several access points. For instance 00:14:BF:97:90 in the top right panel of the figure.*
```{r fig1_6, fig.height=8, fig.width=8}
oldPar = par(mar = c(3.1, 3, 1, 1))
library(lattice)
bwplot(signal ~ factor(angle) | mac, data = offline, 
       subset = posX == 2 & posY == 12, 
       layout = c(2,4))
par(oldPar)
```

## Figure 1.7: Distribution by Signal by Angle for Each Access Point  

```{r, fig.height=12, fig.width=8}
oldPar = par(mar = c(3.1, 3, 1, 1))

densityplot( ~ signal | mac + factor(angle), data = offline,
             subset = posX == 24 & posY == 4, #& mac != "00:0f:a3:39:dd:cd",
             bw = 0.5, plot.points = FALSE)

par(oldPar)
```


```{r}
## Creating composite XY variable
offline$posXY = paste(offline$posX, offline$posY, sep = "-")

## create a dataframe for each combinations of posXY, angle, mac
byLocAngleAP = with(offline, 
                    by(offline, list(posXY, angle, mac), 
                       function(x) x))

## Summarize signal for each posXY, angle, mac
signalSummary = 
  lapply(byLocAngleAP,            
         function(oneLoc) {
           ans = oneLoc[1, ]
           ans$medSignal = median(oneLoc$signal)
           ans$avgSignal = mean(oneLoc$signal)
           ans$num = length(oneLoc$signal)
           ans$sdSignal = sd(oneLoc$signal)
           ans$iqrSignal = IQR(oneLoc$signal)
           ans
           })

offlineSummary = do.call("rbind", signalSummary)  
```

```{r}
## Creating composite XY variable
online$posXY = paste(online$posX, online$posY, sep = "-")

## create a dataframe for each combinations of posXY, angle, mac
byLocAngleAPOnline = with(online, 
                    by(online, list(posXY, angle, mac), 
                       function(x) x))

## Summarize signal for each posXY, angle, mac
signalSummaryOnline = 
  lapply(byLocAngleAPOnline,            
         function(oneLoc) {
           ans = oneLoc[1, ]
           ans$medSignal = median(oneLoc$signal)
           ans$avgSignal = mean(oneLoc$signal)
           ans$num = length(oneLoc$signal)
           ans$sdSignal = sd(oneLoc$signal)
           ans$iqrSignal = IQR(oneLoc$signal)
           ans
           })

#onlineSummary = do.call("rbind", signalSummaryOnline)  
```
## Figure 1.8 SD Signal Stength by Mean Signal Strength
 
```{r}
oldPar = par(mar = c(3.1, 3, 1, 1))

breaks = seq(-90, -30, by = 5)
bwplot(sdSignal ~ cut(avgSignal, breaks = breaks),
       data = offlineSummary, 
       #subset = mac != "00:0f:a3:39:dd:cd",
       xlab = "Mean Signal", ylab = "SD Signal")

par(oldPar)
```
## Figure 1.9 Comparison of Mean and Median Signal Strength  
```{r}
oldPar = par(mar = c(4.1, 4.1, 1, 1))

with(offlineSummary,
     smoothScatter((avgSignal - medSignal) ~ num,
                   xlab = "Number of Observations", 
                   ylab = "mean - median"))
abline(h = 0, col = "#984ea3", lwd = 2)

lo.obj = 
  with(offlineSummary,
       loess(diff ~ num, 
             data = data.frame(diff = (avgSignal - medSignal),
                               num = num)))

lo.obj.pr = predict(lo.obj, newdata = data.frame(num = (70:120)))
lines(x = 70:120, y = lo.obj.pr, col = "#4daf4a", lwd = 2)

par(oldPar)
```
## Figure 1.10  
```{r}
#oneAPAngle = subset(offlineSummary, 
#                    mac == subMacs[1] & angle == 0)


library(fields)
#smoothSS = Tps(oneAPAngle[, c("posX","posY")], 
#               oneAPAngle$avgSignal)

#vizSmooth = predictSurface(smoothSS)

#plot.surface(vizSmooth, type = "C")

#points(oneAPAngle$posX, oneAPAngle$posY, pch=19, cex = 0.5)

surfaceSS = function(data, mac, angle = 45) {
  require(fields)
  oneAPAngle = data[ data$mac == mac & data$angle == angle, ]
  smoothSS = Tps(oneAPAngle[, c("posX","posY")], 
                 oneAPAngle$avgSignal)
  vizSmooth = predictSurface(smoothSS)
  plot.surface(vizSmooth, type = "C", 
               xlab = "", ylab = "", xaxt = "n", yaxt = "n")
  points(oneAPAngle$posX, oneAPAngle$posY, pch=19, cex = 0.5) 
}

parCur = par(mfrow = c(2,2), mar = rep(1, 4))

mapply(surfaceSS, mac = subMacs[ rep(c(2, 1), each = 2) ], 
       angle = rep(c(0, 135), 2),
       data = list(data = offlineSummary))
 
par(parCur)

#offlineSummary = subset(offlineSummary, mac != subMacs[2])

AP = matrix( c( 7.5, 6.3, 7.5, 6.3, 2.5, -.8, 12.8, -2.8,  
                1, 14, 33.5, 9.3,  33.5, 2.8),
            ncol = 2, byrow = TRUE,
            dimnames = list(subMacs, c("x", "y") ))

AP

diffs = offlineSummary[ , c("posX", "posY")] - 
          AP[ offlineSummary$mac, ]

offlineSummary$dist = sqrt(diffs[ , 1]^2 + diffs[ , 2]^2)
```

## Figure 1.11 Signal Stength vs. Distance to Access Point
```{r , fig.height=12, fig.width=8}
oldPar = par(mar = c(3.1, 3.1, 1, 1))
library(lattice)
xyplot(signal ~ dist | factor(mac) + factor(angle), 
       data = offlineSummary, pch = 19, cex = 0.3,
       xlab ="distance")
par(oldPar)
```







