---
title: "Case Study - Real Time Location Systems"
date: "`r Sys.Date()`"
author: "Allison Roderick, Jenna Ford, William Arnost"
output:
  rmdformats::readthedown:
    number_sections: true
    code_folding: hide
    highlight: kate
    toc_depth: 3
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(formatR)

library(tidyverse)
library(magrittr)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


```{r datapull1, echo=FALSE}
# Read lines from offline file
# This creates a character vector for every line in the file
txt = readLines("Data/GeoLoc_offline.final.trace.txt")

#######
# This is taken from the R text book.  
# Basically, read in the line, split it by tokens and then assemble it into a data matrix.
#######
processLine = function(x) {
  # We split the string at ; = and , based on the format of this file
  tokens = strsplit(x, "[;=,]")[[1]]
  
  # Strings with only 10 tokens stop before a mac reading
  if (length(tokens) == 10)
      return(NULL)
  
  tmp = matrix(tokens[ - (1:10) ], , 4, byrow = TRUE)  #build a 4 column matrix mac,signal,scan,type
  cbind(matrix(tokens[c(2, 4, 6:8, 10)], nrow(tmp), 6, #builds a 6 column matrix t,scanMac,x,y,z,angle
        byrow = TRUE), tmp) #binds them together
}

#process the whole datafile, first removing the top comments
lines = txt[ substr(txt, 1, 1) != "#" ]
tmp = lapply(lines, processLine)
offline = as.data.frame(do.call("rbind", tmp),stringsAsFactors = FALSE)

# create a variable with the 7 distinct mac names
subMacs = names(sort(table(offline$mac), decreasing = TRUE))[1:7]


# process the online data file by same method as offline
txt = readLines("Data/GeoLoc_online.final.trace.txt")
lines = txt[ substr(txt, 1, 1) != "#" ]
tmp = lapply(lines, processLine)
online = as.data.frame(do.call("rbind", tmp),stringsAsFactors = FALSE)

# assign variable names to both offline and online data
names(offline) = c("time", "scanMac", "posX", "posY", "posZ",
"orientation", "mac", "signal",
"channel", "type")
names(online) = c("time", "scanMac", "posX", "posY", "posZ",
"orientation", "mac", "signal",
"channel", "type")

# narrow down to only type 3
# (the only other type is type 1, which are ad hoc per the book)
offline = offline[offline$type==3,]
online = online[online['type']==3,]

```

# Introduction, Background, and Methods

This case study explores the data captured by real time location systems (RLTS). The data provided indicates the signal strength, as measured in decibels (dB) by a hand-held device, of various fixed access points. Both the scanning devices and the fixed devices are identified in the data by their media access control (MAC) addresses.

The data is organized as a text file with each line representing a measurement event. An example line and each variable recorded are shown below.

```{r txt}
# this is the first non-comment line in the text file
txt[4]
```

* Time of measurement event `t`
* Scanning device's MAC address `id` 
  + All measurements are taken by the same scanning device, so each line has the same MAC address `00:02:2D:21:0F:33`
* Position measured on x, y, and z axes `pos`
  + The z-axis measurement is always 0
* Orientation of the scanning devices `degree`
* Data regarding the measurement recorded by up to 12 different fixed devices, which are broken into these four different fields:
  + MAC address (the possible addresses are below)
    - Because there are only 6 access points in the physical location this data was recorded at, we will only consider the top 7 addresses by frequency for our analysis (MAC addresses 00:0f:a3:39:e1:c0 and 00:0f:a3:39:dd:cd represent the same access point)
  + Signal strength (measured in dB)
  + Channel
  + Device type
    - We will omit all device types other than 3 as these are ad-hoc measurements beyond the scope of this analysis


```{r unique_macs}
# investigating frequency of measurements by mac address
vals = data.frame(table(mac=offline['mac']))
vals[order(-vals$Freq),]
```

Each measurement event can be uniquely identified by the time, position, and orientation of the scanning device. However, our initial processing of the text file created a "long" dataframe that had multiple rows for each unique event--one row for each of the up to 7 fixed access devices at which the measurement was recorded. We then transposed the data from "long" to "wide", in order to make each row a unique event, with 7 new variables, each representing a fixed access device's measurement.


``` {r datapull2}

# format signal stength as an integer
offline$signal %<>% as.integer
online$signal %<>% as.integer

# limit to top 7 macs
online = online[ online$mac %in% subMacs, ]
offline = offline[ offline$mac %in% subMacs, ]

# remove channel and scanMac as they have no bearing on the analysis
# transpose long to wide so each of the 7 macs has a column
offline_pivot<-select(offline, -c(channel,scanMac)) %>% pivot_wider(names_from = mac,values_from = signal, values_fn = list(signal=mean))
online_pivot<-select(online, -c(channel,scanMac)) %>% pivot_wider(names_from = mac,values_from = signal, values_fn = list(signal=mean))

# check for nas
offline_pivot$nas<-rowSums(is.na(offline_pivot))

# remove nas
offline_pivot = offline_pivot[offline_pivot$nas==0,]

# create categorical variable representing the x-y position 
offline_pivot$posXY = paste(offline_pivot$posX, offline_pivot$posY, sep = "-")
online_pivot$posXY = paste(online_pivot$posX, online_pivot$posY, sep = "-")

```
